---
title: "Basic building blocks of R progrmming"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  basic building blocks of R programming
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.timelimit = 60)
```

## Welcome and grants

Hello!

This is a tutorial adapted from the swirl source codeat [GitHub](https://github.com/swirldev/swirl_courses). 

I have just slightly modified its original content to adapt it for my specific purposes (thanks again to 'swirl' authors for such a great 'free lunch').

## Outline

We will next explore some basic building blocks of the R programming language.

In its simplest form, R can be used as an interactive calculator but it's lik that, obviously!

## How to get help!

If at any point you'd like more information on a particular topic related to R, you can type help.start() at the prompt, which will open a menu of resources  (either within RStudio or your default web browser, depending on your setup).

Alternatively, a simple web search often yields the answer you're looking for.

Anytime you have questions about a particular function, you can access R's built-in help files via the `?` command. For example, if you want more information on the c() function (we'll see next what is about).

Give it a try now, type ?c without the parentheses that normally follow the function name.


```{r help, exercise=TRUE}

```

```{r help-hint}
?c
```

## Is R just a calculator? ...Noope!

In its simplest form, R can be used as an interactive calculator.

Type 5 + 7 and then press button `Run Code`.

```{r sum, exercise=TRUE}

```

```{r sum-hint}
5 + 7
```

R prints the result of 12 by default.

However, R is a programming language and often the reason we use it as opposed to a simple calculator is to automate some process or avoid unnecessary repetition.

In this case, we may want to use our result from above in a second calculation. 
Instead of retyping 5 + 7 every time we need it, we can just create a new variable that stores the result.

The way you assign a value to a variable in R is by using the assignment operator, which is just a 'less than' symbol followed by a 'minus' sign. It looks like this: <-

Think of the assignment operator as an arrow. You are assigning the value on the right side of the arrow to the variable name on the left side of the arrow.

To assign the result of 5 + 7 to a new variable called x, you type `x <- 5 + 7`.
This can be read as 'x gets 5 plus 7'. Give it a try now.

It's important to include a single space on each side of the assignment operator, but do NOT put a space between the `<` and the `-` that form the arrow `<-`.

```{r assignx, exercise=TRUE}

```

```{r assignx-hint}
x <- 5 + 7
```
You have certainly noticed that nothing happened in the "outside" because R did not print the result of 12 this time. When you use the assignment operator, R assumes that you don't want to see the result immediately, but rather that you intend to use the result for something else later on. 

To view the content of the variable `x`, just type `x` and then press button `Run Code`. Try it now.

```{r showx, exercise=TRUE, exercise.setup="assignx-hint"}

```

```{r showx-hint}
x
```

Now, let's imagine to store the result of `x - 3` in a new variable called `y`. Whether you use `x - 3` or `x-3` is just matter of personal preferences: remember it's good habit to include a space on either side of the assignment operator.

```{r assigny, echo=FALSE}
question("How would you do that assignment to the new varaible y?",
  answer("y <- x - 3", correct = TRUE),
  answer("x <- y - 3"),
  answer("y -3")
)
```

```{r answery, echo=FALSE}
question("What do you think is now the value of the new variable  y?",
  answer("-3"),
  answer("x"),
  answer("9", correct = TRUE)
)
```


<!-- - Class: text -->
<!--   Output: Now, let's create a small collection of numbers called a vector. Any object -->
<!--     that contains data is called a data structure and numeric vectors are the simplest -->
<!--     type of data structure in R. In fact, even a single number is considered a vector -->
<!--     of length one. -->

<!-- - Class: cmd_question -->
<!--   Output: The easiest way to create a vector is with the c() function, which stands -->
<!--     for 'concatenate' or 'combine'. To create a vector containing the numbers 1.1, -->
<!--     9, and 3.14, type c(1.1, 9, 3.14). Try it now and store the result in a variable -->
<!--     called z. -->
<!--   CorrectAnswer: z <- c(1.1, 9, 3.14) -->
<!--   AnswerTests: omnitest(correctExpr='z <- c(1.1, 9, 3.14)') -->
<!--   Hint: Inputting z <- c(1.1, 9, 3.14) will assign the vector (1.1, 9, 3.14) to a -->
<!--     new variable called z. Including single spaces after the commas in the vector -->
<!--     is not required, but helps make your code less cluttered and more readable. -->

