---
title: "Vector subsetting"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  missing values

---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.timelimit = 60)
```

# Introduction

In this exercise, we'll see how to extract elements from a vector based on some specific conditions.
For example, we may only be interested in the first 20 elements of a vector, or only the elements that are not NA, or only those that are positive or correspond to a specific variable of interest. By the end of this exercise, you'll hopefully know how to handle each of these scenarios.

This is a tutorial adapted from the swirl source code at [GitHub](https://github.com/swirldev/swirl_courses). 

I have just slightly modified its original content to adapt it for my specific purposes (thanks again to 'swirl' authors for such a great 'free lunch').

# Subsetting

```{r initial_setup, include=FALSE}
set.seed(12345)
x <- sample(c(rnorm(20), rep(NA, 20)))
```

Let's start by using a vector called `x` that contains a random ordering of 20 numbers (from a standard normal distribution) and 20 NAs. Type `x` now to see what it looks like.

```{r x_show, exercise=TRUE, exercise.setup="initial_setup"}

```


```{r q_x}
  question("Can you guess the code to generate the vector `x`?",
    answer("x <- sample(c(rnorm(NA), rep(20)))"),
    answer("x <- sample(c(rnorm(20), rep(20, NA)))"),
    answer("x <- sample(c(rnorm(20), rep(NA, 20)))", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
  )
```

The way you tell R that you want to select some particular elements (i.e. a 'subset') from a vector is by placing an 'index vector' in square brackets immediately following the name of the vector.

For a simple example, try x[1:10] to view the first ten elements of x.

```{r x_subset, exercise=TRUE, exercise.setup="initial_setup"}

```

```{r x_subset-hint}
x[1:10]
```

# Indexing 

Index vectors come in four different flavors:

- logical vectors
- vectors of positive integers
- vectors of negative integers
- vectors of character strings.

## Logical vectors

One common scenario when working with real-world data is that we want to extract all elements of a vector that are not NA (i.e. missing data). Recall that is.na(x) yields a vector of logical values the same length as x, with TRUEs corresponding to NA values in x and FALSEs corresponding to non-NA values in x.

Remember that is.na(x) tells us **where** the `NA`s are in a vector. 
So if we subset x based on that...

```{r q_logical}
  question("What do you think `x[is.na(x)]` will give you?",
    answer("A vector with no NAs"),
    answer("A vector of TRUEs and FALSEs"),
    answer("A vector of length 0"),
    answer(" A vector of all NAs", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

Prove it to yourself by entering the code below.

```{r x_is_na, exercise=TRUE, exercise.setup="initial_setup"}

```

```{r x_is_na-hint}
x[is.na(x)]
```

This is clearly not what we want! In fact, we need to get all elements of a vector that are *NOT* `NA`.
Recall that `!` gives us the negation of a logical expression, so `!is.na(x)` can be read as 'is not NA'. Therefore, if we want to create a vector called y that contains all of the non-NA values from x, we can use it to capture all non-missing values from x. Try it now!

```{r x_is_not_na, exercise=TRUE, exercise.setup="initial_setup"}

```

```{r x_is_not_na-hint}
y<-x[!is.na(x)]
y
```

Now that we've isolated the non-missing values of `x` and put them in `y`, we can subset `y` as we please.

Recall that the expression y > 0 will give us a vector of logical values the same length as y, with TRUEs corresponding to values of y that are greater than zero and FALSEs corresponding to values of y that are less than or equal to zero. 

```{r q_y_subset}
  question("What do you think `y[y > 0]` will give you?",
    answer("A vector of all the negative elements of y"),
    answer("A vector of all NAs"),
    answer("A vector of length 0"),
    answer("A vector of TRUEs and FALSEs"),
    answer("A vector of all the positive elements of y", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

Type `y[y > 0]` to see that we get all of the positive elements of `y`, which are also the positive elements of our original vector `x`.

```{r y_subset_pos, exercise=TRUE, exercise.setup="initial_setup"}

```

You might wonder why we didn't just start with x[x > 0] to isolate the positive elements of x. 
Try that now to see why to see why we didn't use this approach.

```{r x_subset_pos, exercise=TRUE, exercise.setup="initial_setup"}

```

Since NA is not a value, but rather a *placeholder* for an unknown quantity, the expression `NA > 0` evaluates to `NA`. Hence we get a bunch of `NA`s mixed in with our positive numbers when we do this.

By combining our knowledge of logical operators with our new knowledge of subsetting, we could easily do that just in one shot.

```{r q_x_subset_oneshot}
  question("What do you think is the right one?",
    answer("x[!is.na(x) & x != 0]"),
    answer("x[is.na(x) & x > 0]"),
    answer("x[!is.na(x) & x > 0]", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

In this case, we request only values of x that are both non-missing AND greater than zero.

## Positive integers vectors

We have already seen how to subset just the first ten values of x using x[1:10]. In this case, we're providing a vector of positive integers inside of the square brackets, which tells R to return only the elements of x numbered 1 through 10.

Many programming languages use what's called 'zero-based indexing', which means that the first element of a vector is considered element `0`. R uses 'one-based indexing', which (you guessed it!) means the first element of a vector is considered element `1`.


```{r q_x_subset_int}
  question("Can you figure out how we'd subset the 3rd, 5th, and 7th elements of x?",
    answer("x[3:7]"),
    answer("x[3, 5, 7]", message="missing something...?"),
    answer("x[c(3, 5, 7)]", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

It's important that when using integer vectors to subset our vector x, we stick with the set of indexes {1, 2, ..., 40} since x only has 40 elements. 

What happens if we ask for the zeroth element of x (i.e. x[0])? 
Give it a try.

```{r x_0, exercise=TRUE, exercise.setup="initial_setup"}

```

```{r x_0-hint"}
# remember what we said about the 'one-based indexing' in R, meaning that the zeroth element of a vector is not defined
```

As you might expect, we get nothing useful. Unfortunately, R doesn't prevent us from doing this. 

What if we ask for the 3000th element of x? Try it out.

```{r x_3000, exercise=TRUE, exercise.setup="initial_setup"}
x[3000]
```

```{r x_3000-hint}
# we are simply asking for an elemnet which does not exist at all
```

Again, nothing useful, but R doesn't prevent us from asking for it. 
This should be a cautionary tale. 
You should always make sure that what you are asking for is within the bounds of the vector you're working with.

## Negative integers vectors

What if we're interested in all elements of x EXCEPT the 2nd and 10th?
It would be pretty tedious to construct a vector containing all numbers 1 through 40 EXCEPT 2 and 10.
Luckily, R accepts negative integer indexes. 

```{r q_x_except}
  question("Whereas x[c(2, 10)] gives us ONLY the 2nd and 10th elements of x, can you figure out what gives us all elements of x EXCEPT for the 2nd and 10 elements?",
    answer("x[-2, - 10]", message= "missing something...?"),
    answer("x[-c(2, 10)]", correct=TRUE, message="this is a shorthand"),
    answer("x[c(-2, -10)]", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

A shorthand way of specifying multiple negative numbers is to put the negative sign out in front of the vector of positive numbers. Therefore `x[-c(2, 10)]` get the exact same result as `x[c(-2, -10)]`.
This could be a usefull time saver and error avoiding strategy when specifying many negative indexes.

## Character string vectors

So far, we've covered three types of index vectors -- logical, positive integer, and negative integer. 
The only remaining type requires us to introduce the concept of 'named' elements.


```{r vect, include=FALSE}
vect <- c(foo = 11, bar = 2, norf = NA)
```

Create a numeric vector with three named elements by using `vect <- c(foo = 11, bar = 2, norf = NA)` and then print it.

```{r vect_print, exercise=TRUE, exercise.setup="vect"}

```

```{r vect_print-hint}
vect <- c(foo = 11, bar = 2, norf = NA)
vect
```

When we print `vect` to the console, we see that each element has a name.
We can also get the names of `vect` by passing it as an argument to the function `names()`.
Give that a try.

```{r names_vect, exercise=TRUE, exercise.setup="vect"}

```

```{r names_vect-hint}
names(vect)
```

Alternatively, we can create an (*ordinary*) unnamed vector, let's say `vect2` that contains `c(11, 2, NA)`. 
Do that now.

```{r vect2, exercise=TRUE}

```

```{r vect2-hint}
vect2 <- c(11, 2, NA)
```

Then, we can add the `names` attribute to `vect2` after the fact with: `names(vect2) <- c("foo", "bar", "norf")`.

```{r q_vect2_names}
  question("Do you think that is correct?",
    answer("no"),
    answer("yes", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

Now, let's check that `vect` and `vect2` are the same 8identical objects) by passing them as arguments to the `identical()` function. The identical() function tells us if its first two arguments are, well, identical!

```{r  cfr, include=FALSE}
vect <- c(foo = 11, bar = 2, norf = NA)
vect2 <- c(11, 2, NA)
```

```{r cfr_vec_vect2, exercise=TRUE, exercise.setup="cfr"}

```

```{r cfr_vec_vect2-hint}
identical(vect, vect2)
```

Indeed, `vect` and `vect2` are identical named vectors.

Now, back to the matter of subsetting a vector by named elements. 
If we want the element named `bar` (i.e. the second element of vect), which command would get us that?

```{r q_subset_2}
  question("Which of the following commands do you think would give us the 2nd element of vect?",
    answer("vect[bar]"),
    answer("vect['2']"),
    answer("vect['bar']", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

Likewise, we can specify a vector of names with `vect[c("foo", "bar")]`.

```{r q_subset_foo_bar}
  question("What do you expect is returning this command?",
    answer("all elements of `vect`"),
    answer("only the element of `vect` named `norf`"),
    answer("only the elements of `vect` named `foo` and `bar`", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
```

Now we know all four methods of subsetting data from vectors. 
Different approaches are best in different scenarios and when in doubt, try it out to see what happens!
